Adaptive evolution with aneuploidy and mutation

### 2021-12-29 ms/ms.tex
revised ABC-SMC description

%%% Table: multi-locus model
\begin{table}[h]
\centering
\caption{\textbf{Multi-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype} $i$ & $2n$ & $2n+1$ & $2n+1^{*k}$ & $2n^{*k}$ \\
\hline
\emph{Fitness} $w_i$ & $1$ & $1-c+b$ & $(1-c)\prod_{j=1}^k(1+s_i)+b$ & $\prod_{j=1}^k(1+s_i)$               
\end{tabular}
\label{table:multi-locus}
\caption*{
$k$ is the number of accumulated beneficial mutations;
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
% TODO check if b > c in code
}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We use the results of evolutionary experiments reported by \citet{Yona2012}.
In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}. Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier). From unpublished results, aneuploidy did not fix before at least 200 generations elapsed.
The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350.
%Thus, the empirical evidence can be summarized by $\vec{X}=\Big(200 \le T_i(2n+1) \le 450,\, i=1,2,3,4;\ T_1(2n^*)=1700;\ T_2(2n^*)=2350\Big)$ where $T_i(g)$ is the fixation time of genotype $g$ in population $i$.

\paragraph{Likelihood function.}

Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
We overcome this problem by approximating the likelihood using simulations. We simulate 1,000 experiments per parameter vector $\theta = (\mu, \tau, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$. We then compute the approximate likelihood,
\begin{equation}\begin{aligned}
\label{eq:heatstress-likelihood}
\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
	\Big[1 - \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
	\Big]\;,
\end{aligned}\end{equation}
where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$. For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.

%Denote by $A_{t,T}$ the fixation of aneuploidy by generation $t$ conditioned on no fixation up to generation $T$; by $B_{t}$ the extinction of aneuploidy by generation $t$; and by $C_{t,T}$ the extinction of aneuploidy by generation $t$ conditioned on no extinction by generation $T$.
%The likelihood function for parameter vector $\theta$ conditioned on the observed data $\vec{X}$ from the heat-stress experiment is
%\begin{multline} \label{eq:heatstress-likelihood}
%P^4(A_{450,200}) \cdot \Big(1 - P^4(\lnot{B_{1700}} | A_{450,200} ) - P^4(\lnot{C_{2350, 1700}}| A_{450,200}) + P^4(\lnot{B_{1700}} \land \lnot{C_{2350, 1700}}| A_{450,200})\Big)	\;,	 % TODO fix L* (YR)
%\end{multline}
%where $P(X)$ is the probability of event $X$ and $\lnot$ means \emph{not}.

For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), 
%denote by $M_{t}$ the fixation of \eumt\ by generation $t$, and by $Q_{t,T}$ the fixation of \eumt\ by generation $t$ conditioned on no fixation at generation $T$.
the likelihood is similarly approximated by
% of a model without aneuploidy with parameter vector $\theta$ for the heat-stress experiment is
\begin{equation}\begin{aligned}
\label{eq:heatstress-noaneuploidy-likelihood}
\likelihood_{!}(\theta) = &\ 
	1 - 
	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
\;.
\end{aligned}\end{equation}

%\begin{multline} \label{eq:heatstress-noaneuploidy-likelihood}
%\likelihood_{!}(\theta) = 
%\Big(1 - P^4(\lnot{M_{1700}}) - P^4(\lnot{Q_{2350, 1700}}) + P^4(\lnot{M_{1700}} \land \lnot{Q_{2350, 1700}})\Big)	\;,	
%\end{multline}

@@

\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC} Python package~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
This approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters. It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be directly computed. It is therefore suitable to our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.

Briefly, the ABS-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Syga2021}. 
In iteration $t$ of the algorithm, we continue to sample a parameter vector $\theta$, also called \emph{particle}, and simulate observations $\tilde{\vec X}$, until $n_t$ particles are accepted. 
A particle $\theta$ is accepted if its approximate likelihood is high enough, $\likelihood(\theta)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted. % TODO something about how the weight is computed and used?
The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
The proposed particles are sampled from a proposal distribution: for $t=0$, this is the prior distribution, and for $t>0$ the proposal distribution is constructed using a multivariate normal kernel with adaptive covariance matrix from the particles accepted in the previous iteration. % TODO ref - Klinger? Syga? Sisson?
The number of particles, $n_t$, is adapted at every iteration $t$ using the \emph{adaptive population strategy}~\citep{Klinger2017}.

@@

We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$. In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$. 
%TODO please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
We tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$.
In both of these uninformative priors, the fitness estimates were much too high: for the first prior the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second prior the estimates were $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$. Such high fitness values are unreasonable. Moreover the WAIC for the models with these priors were 0.57 and 1.9, respectively, whereas the WAIC with the informative prior was 0.27. 
% TODO if you need to remove this sentence: Also, with the uninformative prior for all fitness values, the fixation time of \anwt is about 16 on average, much lower than the (unpublished) experimental result, which is 200-300 generations.

We tried to use additional growth curves. We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$. The same prior was used for $w_{2n*}/w_{2n+1*}$. This prior resulted in WAIC 2.8, compared to 0.27 with the above informative prior.
We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} to estimate $1-c$. This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values. The prior for $b$ was taken the same as for $c$. This prior did not work at all, as no parameter sets were found with an approximate likelihood greater than zero.
% TODO it make sense because bla bla (to explain)

% TODO : Actually I assume in code that  w(2n*)>w(2n+1*)>w(2n+1)>w(2n) just for speeding up the simulations (I don't check some parameters combinations in this way). But I checked also that when I don't adding this constrain the result was the same. However I dind't add reverse direction in this case.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\autoref{fig:posterior}):
mutation rate, $\mu=3.06\substack{+1.229 \\ -2.84}\times10^{-6}$,
aneuploidy rate, $\delta=1.61\substack{+1.122 \\ -0.197}\times10^{-3}$,
and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
$w_{2n+1*}=1.025\substack{+0.001 \\ -0.001}$,
$w_{2n*}=1.028\substack{+0.001 \\ -0.002}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
% describe b and c
This estimated aneuploidy rate agrees with previous estimates. % TODO refs
The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
%We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes. 
% TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?

@@

\paragraph{Model checking and comparison.}
The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 59\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\autoref{fig:2n*-fixation}) and \anwt\ fixed in roughly 300 generations on average (green lines in \autoref{fig:ppc-plot}), which agrees with experimental results.
% TODO replace "roughly 300" with the exact number.
% TODO this fig shows fixation of 2n*, what about fixation of 2n+1?
% TODO merge fig 3 and fig:likelihood (both panels of fig:likelihood) 
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\autoref{fig:ppc-plot}). % TODO move somewhere else?
Furthermore, sensitivity analysis shows that changing the parameter values from the MAP estimate reduces the model fit to the experimental results~(\autoref{fig:sensitivity}).

However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figures~\ref{fig:2n*-fixation} and \ref{fig:likelihood}). The estimated parameter values without aneuploidy are:
mutation rate, $\mu=1.251\substack{+0.146 \\ -0.177}\times10^{-8}$,
%2n+1 fitness - $1.01\substack{+0.001 \\ -0.002}$,
%2n+1* fitness - $1.011\substack{+0.001 \\ -0.001}$,
%TODO why do you have fitness of aneuploidy when there is no aneuploidy?
fitness of mutant, $w_{2n^*}=1.011\substack{+0.001 \\ -0.001}$. 
We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy. Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental obvservations (\autoref{fig:2n*-fixation}). Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
% TODO WAIC?

% Fig single-locus model with and without aneuploidy
\begin{figure}[b!]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
      \caption{
        \textbf{A. With aneuploidy}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
      \label{fig:likelihood-with-aneuploidy}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
  	  \caption{
        \textbf{B. Without aneuploidy}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/likelihood-noaneuploidy-plot.pdf}      
      \label{fig:likelihood-without-aneuploidy}
  \end{subfigure}
  \caption{
  % TODO change x label - "Time to fixation of 2n*"
  % plot both lines on one panel with two colors (blue and red for with and without? choose nice color, not the default 'r' and 'b'!), make the without in dashed.
    \textbf{Single-locus model fit with and without aneuploidy.} Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
    \textbf{(B)} Simulations parameters are from the maximum a posterior estimation of the single-locus model \emph{without} aneuploidy fit. 
     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.86 and 0.75 with and without aneuploidy. % TODO likelihood ratio test? bayes factor? no sense in just giving the likelihoods.
  }
  \label{fig:likelihood}
\end{figure}

\begin{figure}[h!]
  \centering
  % TODO add legend
\includegraphics[width=0.6\textwidth]{../figures/ppc-plot.pdf}
  \caption{
  \textbf{Single-locus model posterior genotype frequency dynamics.} 
  The posterior predicted frequency of the four genotypes. Each line is the average of 10,000 simulations using parameters drawn from the posterior distribution (\autoref{fig:posterior}).
  }
  \label{fig:ppc-plot}
\end{figure}

\begin{figure}[h!]
  \centering
\includegraphics[width=0.6\textwidth]{../figures/fixation-plot.pdf}
  \caption{
  \textbf{Adaptation time cannot be explained without aneuploidy.}
In blue, the distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: in orange, a model with the same parameter values except $\delta=0$; in green, a model fitted to the data assuming $\delta=0$.
%, where parameters are taken from the median of the posterior of the inference. % why madian and not MAP?
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 200. Thus, the blue distribution demonstrates a better fit compared to the green, and the yellow demonstrates a very poor fit.
The last bin contains all the simulations with time equal or greater than 3000.
}
% TODO add legend
% TODO change y label to % simulations
% change x label to Time to fixation of 2n*
% makes vertical line black
% add line for t=200?
% does x axis start at zero? can you add 0 to x ticks?
  \label{fig:2n*-fixation}
\end{figure}

@@

% Fig Curveball 39deg
% generated with growth_curves.ipynb
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/evo39_fitness_39deg.pdf}
    \caption{
    \textbf{Fitness estimation from \SI{39}{\celsius}.}
    } 
    \label{fig:growth-curves-39deg}
\end{figure}
% TODO add caption, combine to one figure with the previous one

>>>>

%%% Table: multi-locus model
\begin{table}[h]
\centering
\caption{\textbf{Multi-locus model fitness values.}}
\begin{tabular}{lllll}
\emph{Genotype}, $g$ & $2n$ & $2n+1$ & $2n+1^{*k}$ & $2n^{*k}$ \\
\hline
\emph{Fitness}, $w_g$ & $1$ & $1-c+b$ & $(1-c)\prod_{j=1}^k(1+s_i)+b$ & $\prod_{j=1}^k(1+s_i)$               
\end{tabular}
\label{table:multi-locus}
\caption*{
$k$ is the number of accumulated beneficial mutations;
$s \ge 0$ is the selection coefficient of a beneficial mutation;
$0 \le c \le 1$ is the fitness cost of aneuploidy;
and $b \ge c$ is the selection coefficient, or fitness benefit, of aneuploidy.
}
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We use the results of evolutionary experiments reported by \citet{Yona2012}.
In their heat-stress experiment, four populations of \yeast evolved under \SI{39}{\celsius}. Aneuploidy fixed in all four population in the first 450 generations (hereafter, fixation or elimination of a genotype \emph{by generation $t$} means that more than 95\% or less than 5\% of the population carry the genotype at generation $t$, and possibly earlier). From unpublished results, aneuploidy did not fix before at least 200 generations elapsed.
The experiment continued with two populations, in which aneuploidy was eliminated by generation 1,700 and 2,350.

\paragraph{Likelihood function.} 
% TODO please read!
Because our model, just like the Wright-Fisher model, is non-linear and stochastic, computing the distribution of fixation time $T(g)$ of genotype $g$ for use in the likelihood function is intractable (it is even hard to use a diffusion-equation approximation due to the model having multiple genotypes, rather than just two).
We overcome this problem by approximating the likelihood using simulations. We simulate 1,000 experiments per parameter vector $\theta = (\mu, \tau, \delta, s, b, c)$, resulting in a set of simulated observations $\tilde{\vec X} = \{\tilde{X}_i\}_{i=1}^{1000}$. We then compute the approximate likelihood,
\begin{equation}\begin{aligned}
\label{eq:heatstress-likelihood}
\likelihood(\theta) = &\ P^4(200 \le T(2n+1) \le 450) \cdot 
	\Big[1 - \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)- \\
	&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big)+ \\
	&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\} \mid 200 \le T(2n+1) \le 450\big) 
	\Big]\;,
\end{aligned}\end{equation}
where $!\{\ldots\}$ is the "logical not" operator, $P^4(\ldots)$ is the 4th power of $P(\ldots)$, and all probabilities $P_{\tilde{\vec X}}(\ldots)$ are approximated from the results of the simulations $\tilde{\vec X}$. For example, $P_{\tilde{\vec X}}\big(!\{T(2n^*)<1700\} \mid 200 \le T(2n+1) \le 450\big)$ is approximated by taking simulations in which \anwt\; fixed before generation 450 but not before generation 200, and computing the fraction of such simulations in which \eumt\; did not fix by generation 1,700, and hence aneuploidy did not extinct before generation 1,700.

For a model without aneuploidy (that is, when the aneuploidy rate is fixed at zero, $\delta=0$), 
the likelihood is similarly approximated by
\begin{equation}\begin{aligned}
\label{eq:heatstress-noaneuploidy-likelihood}
\likelihood_{!}(\theta) = &\ 
	1 - 
	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\}\big) - \\
&	P_{\tilde{\vec X}}^4\big(!\{1700 < T(2n^*) < 2350\}\big) + \\
&	P_{\tilde{\vec X}}^4\big(!\{T(2n^*)<1700\} \land !\{1700 < T(2n^*) < 2350\}\big)
\;.
\end{aligned}\end{equation}

@@

\paragraph{Parameter inference.} To infer model parameters, we use approximate Bayesian computation with a sequential Monte-Carlo scheme, or ABC-SMC~\citep{Sisson2009}, implemented in the \texttt{pyABC} Python package~\citep[\href{https://pyabc.readthedocs.io}{pyabc.readthedocs.io}]{Klinger2018}.
This approach uses numerical stochastic simulations of the model to infer a posterior distribution over the model parameters. It is a method of likelihood-free, simulation-based inference~\citep{Cranmer2020}, that is, for estimating a posterior distribution when a likelihood function cannot be directly computed. It is therefore suitable in our case, in which the likelihood function can only be approximated from simulations, and cannot be directly computed.

%% YR revise
The ABS-SMC algorithm employs sequential importance sampling over multiple iterations~\citep{Toni2009, Klinger2017, Syga2021}.
In iteration $t$ of the algorithm, a set of parameter vectors, $\{\theta_{i,t}\}_{i=1}^{n_t}$, also called \emph{particles}, are constructed in the following way.
A proposal particle, $\theta^*$, is sampled from a proposal distribution, and is either accepted or rejected, until $n_t$ particles are accepted.
The number of particles, $n_t$, is adapted at every iteration $t$ using the adaptive population strategy.
For $t=0$, the proposal particle is sampled from the prior distribution, $p(\theta)$.
For $t>0$, the proposal particle is sampled from the particles accepted in the previous iteration, $\{\theta_{i,t-1}\}_{i=1}^{n_{t-1}}$, each with a probability relative to its weight $w_{t-1}(\theta_{i,t-1})$ (see below). The proposal particle is then perturbed using a kernel perturbation kernel, $K_t(\theta^* \mid \theta)$ where $\theta$ is the sample from the previous iteration.
Then, a set of synthetic observations $\tilde{\vec X}^*$ is simulated, and the proposal particle $\theta^*$ is accepted if its approximate likelihood (\autoref{eq:heatstress-likelihood}) is high enough, $\likelihood(\theta^*)>1-\epsilon_t$ (or more commonly, if $1-\likelihood(\theta^*) < \epsilon_t$), where $\epsilon_t>0$ is the \emph{acceptance threshold}, as higher values of $\epsilon_t$ allow more particles to be accepted. 
The acceptance threshold $\epsilon_t$ is chosen as the median of the $1-\likelihood(\theta)$ of the particles accepted in the previous iteration, $t-1$, and $\epsilon_0=0.01$. 
For each accepted particle $\theta_{i,t}$ a weight $w_t(\theta_{i,t})$ is assigned: for $t=0$, $w_0(\theta_{i,0})=1$, and for $t>0$, 
$w_t(\theta_{i,t}) = p(\theta_{i,t}) / \sum_{i=1}^{n_{t-1}}{w_{t-1}(\theta_{i,t-1}) K_t(\theta_{i,t}, \theta_{i,t-1})}$, where $p(\theta)$ is the prior density of $\theta$ and $K_t(\theta', \theta)$ is the probability of a perturbation from $\theta$ to $\theta'$.
$K_t(\theta' \mid \theta)$ is a multivariate normal distribution. % TODO construction of K_T from \theta_{i,t-1}

@@

We used growth curves of \euwt\ and \anwt\ in \SI{39}{\celsius} to estimate a prior distribution for $w_{2n+1}$. In lieu of a better prior, we used the same prior for $w_{2n+1^*}$ and $w_{2n^*}$.

Compared to other priors we tested, this prior produced lower WAIC, better posterior prediction plots, and more stable parameter estimates, as follows.
%In both of these uninformative priors, the fitness estimates were much too high: for the first prior the median estimates were  $w_{2n+1}=2.767$, $w_{2n+1^*}=5.79$, $w_{2n^*}=5.827$ and for the second prior the estimates were $w_{2n+1}=1.046$, $w_{2n+1^*}=4.642$, $w_{2n^*}=4.684$. Such high fitness values are unreasonable. Moreover the WAIC for the models with these priors were 0.57 and 1.9, respectively, whereas the WAIC with the informative prior was 0.27. 
% TODO if you need to remove this sentence: Also, with the uninformative prior for all fitness values, the fixation time of \anwt is about 16 on average, much lower than the (unpublished) experimental result, which is 200-300 generations.

We tried to use additional growth curves. We used growth curves of \eumt\; (\emph{refined} strain from \citet{Yona2012}) and \anwt\; in \SI{39}{\celsius} to estimate $w_{2n*}/w_{2n+1}$. The same prior was used for $w_{2n*}/w_{2n+1*}$. This prior resulted in WAIC 0.32, compared to 0.27 with the above informative prior.
% TODO what are the results ?

We also tried to use growth curves of \anwt\; and \euwt\; in \SI{30}{\celsius} to estimate $1-c$. This estimation assumes that the cost of aneuploidy is the same in \SI{39}{\celsius} and \SI{30}{\celsius}; this might be incorrect, but we only assumed this to generate a prior distribution for the fitness values. The prior for $b$ was taken the same as for $c$. This prior did not converge: no parameter sets were found with an approximate likelihood greater than zero.
% TODO it make sense because bla bla (to explain)
Finally, we also tested a uninformative uniform prior with $\mathit{U}(1,6)$, for (i) all $w_{2n+1}$, $w_{2n+1^*}$, $w_{2n^*}$ and (ii) only for $w_{2n+1^*}$, $w_{2n^*}$, using the above informative prior for $w_{2n+1}$. In both cases the algorithm failed to converge.

% TODO : Actually I assume in code that  w(2n*)>w(2n+1*)>w(2n+1)>w(2n) just for speeding up the simulations (I don't check some parameters combinations in this way). But I checked also that when I don't adding this constrain the result was the same. However I didn't add reverse direction in this case.

@@

\paragraph{Parameter estimation.} 
We used ABC-SMC to infer the posterior distribution of the parameters of the single-locus model.
The MAP (maximum a posteriori) and 50\% HDI (highest density interval) of the parameters posterior are~(\autoref{fig:posterior}):
mutation rate, $\mu=3.06\substack{+1.229 \\ -2.84}\times10^{-6}$,
aneuploidy rate, $\delta=1.61\substack{+1.122 \\ -0.197}\times10^{-3}$,
and fitness values, $w_{2n+1}=1.022\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.025\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$, all relative to the fitness of \euwt, which is set to $w_{2n}=1$.
% describe b and c
This estimated aneuploidy rate agrees with previous estimates. % TODO refs
The mutation rate corresponds to a mutation target size of $10^{4}$, assuming the mutation rate per base pair is roughly $2\cdot10^{-10}$~\citep{Zhu2014}. 
%We can see from the posterior distribution~(\autoref{fig:posterior}) the correlation between fitness of different genotypes. 
% TODO say/plot something about the correlation: which fitness values are correlated? how strong is it? why does it occur?

We also tried to inference parameters of the single-locus model with different $\tau$: 1, 2, 5, 10 and 100. It was seen that for 1, 2, 5 the posterior was not significantly different from the model with $\tau=33/32$, while for $\tau=10$ the estimated $\mu\ $ was a little lower, $1.654\substack{+0.879 \\ -1.775}\times10^{-6}$, and for $\tau=100$ the estimated $\mu\ $ was already significantly lower and $\delta\ $ had higher variability. The MAP and 50\% HDI for $\tau=100$:
$\mu=5.538\substack{+1.611 \\ -5.311}\times10^{-7}$,
$\delta=8.292\substack{+8.988 \\ -4.96}\times10^{-4}$,
$w_{2n+1}=1.023\substack{+0.001 \\ -0.002}$,
$w_{2n+1^*}=1.024\substack{+0.001 \\ -0.001}$,
$w_{2n^*}=1.028\substack{+0.001 \\ -0.002}$.
%TODO: show joint distributions comparison plot (20-diffk.ipynb). WAIC?

% TODO use tau=1 as the default; we also check tau>1 - 33/32, 2, 5, 10, 100.
% TODO support figure of joint posterior delta vs mu per tau=1, 33/32, 2, 5, 10

@@

\paragraph{Model checking and comparison.}
The single-locus model fits the data well: in simulations using the best-fit parameters (MAP estimate) \eumt\ fixed in 59\% of simulations by generation 1,700 and in 100\% of simulations by generation 2,350~(\autoref{fig:2n*-fixation}) and \anwt\ fixed in roughly 300 generations on average (green lines in \autoref{fig:ppc-plot}), which agrees with experimental results.
% TODO replace "roughly 300" with the exact number.
Interestingly, the genotype frequency dynamics in these simulations demonstrate that \anmt\ never reaches substantial frequency (\autoref{fig:ppc-plot}). % TODO move somewhere else?
Furthermore, sensitivity analysis shows that changing the parameter values from the MAP estimate reduces the model fit to the experimental results~(\autoref{fig:sensitivity}).

However, a model without aneuploidy where the aneuploidy rate is fixed at zero, $\delta=0$, cannot explain the experimental observations~(Figures~\ref{fig:2n*-fixation} and \ref{fig:likelihood}). The estimated parameter values without aneuploidy are:
mutation rate, $\mu=1.251\substack{+0.146 \\ -0.177}\times10^{-8}$,
fitness of mutant, $w_{2n^*}=1.011\substack{+0.001 \\ -0.001}$. 
We can see that the mutation rate estimate in this model is much lower than in the model with aneuploidy. Higher mutation rate would cause quicker fixation of \eumt\ than in the experimental observations (\autoref{fig:2n*-fixation}). Thus, even disregarding the appearance of trisomic cells in the experiment, the evidence supports the inclusion of aneuploidy in the model.
% TODO WAIC?

% Fig single-locus model with and without aneuploidy
\begin{figure}[b!]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
      \caption{
        \textbf{A. With aneuploidy}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/likelihood-plot.pdf}      
      \label{fig:likelihood-with-aneuploidy}
  \end{subfigure}

  \begin{subfigure}[t]{0.45\textwidth}
  	  \caption{
        \textbf{B. Without aneuploidy}
      }
      \centering
      \includegraphics[height=1.8in]{../figures/likelihood-noaneuploidy-plot.pdf}      
      \label{fig:likelihood-without-aneuploidy}
  \end{subfigure}

  \caption{
  % TODO change x label - "Time to fixation of 2n*"
  % plot both lines on one panel with two colors (blue and red for with and without? choose nice color, not the default 'r' and 'b'!), make the without in dashed.
  % place 82% a bit higher
  % xticks - replace 500, 1000,1500,2000,2500 with 450, 1700, 2350
    \textbf{Single-locus model fit with and without aneuploidy.} Cumulative distribution of the time to fixation of \eumt\ in 10,000 simulations using the MAP estimate with and without aneuploidy (i.e., $\delta=0$ is fixed).
     The MAP likelihood (\autoref{eq:heatstress-likelihood}) is 0.86 and 0.75 with and without aneuploidy.
	}
  \label{fig:likelihood}
\end{figure}

% TODO merge previous and next figure to one figure with 2 panels

\begin{figure}[h!]
  \centering
\includegraphics[width=0.6\textwidth]{../figures/fixation-plot.pdf}
  \caption{
  \textbf{Adaptation time cannot be explained without aneuploidy.}
The distribution of adaptation time (fixation of \eumt) in $10,000$ simulations of the single-locus model with aneuploidy (blue; MAP parameters) compared to two models without aneuploidy: a model with the same parameter values except $\delta=0$ (orange); and a model fitted to the data assuming $\delta=0$ (green).
In the experiment by \citet{Yona2012}, one population lost aneuploidy by generation 1,700 and the other by generation 2,350 (dashed lines) but not before generation 200. Thus, the blue distribution demonstrates a better fit compared to the green, and the yellow demonstrates a very poor fit.
The last bin contains all the simulations with time equal or greater than 3,000.
}
% TODO add legend
% TODO change y label to % simulations
% change x label to Time to fixation of 2n*
% makes vertical line black
% add line for t=200?
% does x axis start at zero? can you add 0 to x ticks?
  \label{fig:2n*-fixation}
\end{figure}

\begin{figure}[h!]
  \centering
\includegraphics[width=0.6\textwidth]{../figures/ppc-plot.pdf}
  \caption{
  \textbf{Single-locus model posterior genotype frequency dynamics.} 
  The posterior predicted frequency of the four genotypes. Each line is the average of 10,000 simulations using parameters drawn from the posterior distribution (\autoref{fig:posterior}).
  }
  \label{fig:ppc-plot}
  % TODO add legend
  % xticks - replace 500, 1000,1500,2000,2500 with 450, 1700, 2350
\end{figure}

@@

% Fig Curveball 39deg
% generated with growth_curves.ipynb
\begin{figure}[h]
    \centering
	\includegraphics[width=0.5\textwidth]{../figures/evo39_fitness_39deg.pdf}
    \caption{
    \textbf{Fitness estimation from \SI{39}{\celsius}.}
    } 
    \label{fig:growth-curves-39deg}
\end{figure}
% TODO add caption

%TODO combine to one figure with the previous one and with refined
%TODO please report here the point estimate and 95% CI for the three distributions, I only have for 1-c+b: 0.9863 (95% CI 0.9790 - 0.9939).
